---
title: "Health risks and economic consequences of different weather events"
author: "Widar J"
date: "29 mars 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
sessionInfo()
```

##Synopsis

The goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. 

Specifically, we would like to address the following questions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

##Data Processing

The zipped data file is downloaded from: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2

```{r}

# Episode   -> Events (EVTYPE) 
#               (Thunderstorm Wind, Hail, Tornado and Flood) +++

# Magnitude -> Wind speed (knots)/ Hail size (inches)
# property and crop damage
# fatalities, injuries, and damage amounts

# The fatalities, injuries, and damage amounts appearing in tropical cyclone events are
# attributed only to wind damage experienced in the coastal counties/parishes listed. Other
# tropical cyclone related events such as tornadoes and flooding are listed within their
# separate event types.

# Read data

library(dplyr)

if (!exists(dfStorm))
{
  dfStorm <- read.csv("repdata_data_StormData.csv.bz2")
}

# Find total fatalities per event

dfStormFatalities <- subset(dfStorm, dfStorm$FATALITIES>0)


dfStormFatalitiesTotal <- dfStormFatalities %>%
  mutate(event = EVTYPE) %>%
  group_by(event) %>% # 
  summarise(total_fatalities=sum(FATALITIES)) #%>%  
  
dfStormFatalitiesTotalOrdered <- dfStormFatalitiesTotal[order(dfStormFatalitiesTotal$total_fatalities),]

dfStormFatalitiesTotalOrdered2 <- dfStormFatalitiesTotalOrdered[158:168,] # select the last 10 columns with the highest fatalities

% Find total expenses per event

dfStormExpenses <- subset(dfStorm, dfStorm$PROPDMG>0 | dfStorm$CROPDMG>0, select=c("EVTYPE", "PROPDMG", "PROPDMGEXP", "CROPDMG", "CROPDMGEXP"))

# propdmg, propdmgexp, cropdmg, cropdmgexp
# K = 1000, M = 1000000, B = 1000000000
#K <- 1000

#df$X <- ifelse(df$ACCURACY == 1, ifelse(df$Feedback == 1, "correct", ifelse(df$Feedback == 2, "unknown", "incorrect")), ifelse(df$Feedback == 1, "correct", ifelse(df$Feedback == 2, "unknown", "incorrect")))


dfStormExpenses$prop_exp <- ifelse(dfStormExpenses$PROPDMG > 0, ifelse(dfStormExpenses$PROPDMGEXP == "K", dfStormExpenses$PROPDMG*1000, ifelse(dfStormExpenses$PROPDMGEXP == "M", dfStormExpenses$PROPDMG*1000000, dfStormExpenses$PROPDMG*1000000000)), ifelse(dfStormExpenses$PROPDMGEXP == "K", dfStormExpenses$PROPDMG*1000, ifelse(dfStormExpenses$PROPDMGEXP == "M", dfStormExpenses$PROPDMG*1000000, dfStormExpenses$PROPDMG*1000000000)))

dfStormExpenses$crop_exp <- ifelse(dfStormExpenses$CROPDMG > 0, ifelse(dfStormExpenses$CROPDMGEXP == "K", dfStormExpenses$CROPDMG*1000, ifelse(dfStormExpenses$CROPDMGEXP == "M", dfStormExpenses$CROPDMG*1000000, dfStormExpenses$CROPDMG*1000000000)), ifelse(dfStormExpenses$CROPDMGEXP == "K", dfStormExpenses$CROPDMG*1000, ifelse(dfStormExpenses$CROPDMGEXP == "M", dfStormExpenses$CROPDMG*1000000, dfStormExpenses$CROPDMG*1000000000)))

dfStormExpenses$total_exp <- dfStormExpenses$prop_exp + dfStormExpenses$crop_exp


dfStormExpensesTotal <- dfStormExpenses %>%
  mutate(event = EVTYPE) %>%
  group_by(event) %>% # 
  summarise(total_expenses=sum(total_exp)) #%>%  
  
dfStormExpensesTotalOrdered <- dfStormExpensesTotal[order(dfStormExpensesTotal$total_expenses),]

dfStormExpensesTotalOrdered2 <- dfStormExpensesTotalOrdered[422:431,] # select the last 10 columns with the highest fatalities


```





## Results



```{r}

library(ggplot2)
g <- ggplot(dfStormFatalitiesTotalOrdered2, aes(event,total_fatalities))
# Number of cars in each class:
g + aes(x=reorder(event,total_fatalities,sum),y=total_fatalities,label=total_fatalities,fill=event) +
geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(title = "Events most harmful to population health") +
  xlab("Events") + ylab("Total fatalities")


g <- ggplot(dfStormExpensesTotalOrdered2, aes(event,total_expenses))
# Number of cars in each class:
g + aes(x=reorder(event,total_expenses,sum),y=total_expenses,label=total_expenses,fill=event) +
geom_col() + theme(axis.text.x = element_text(angle = 90, hjust = 1)) + labs(title = "Events with greatest economic consequences") +
  xlab("Events") + ylab("Total expenses")


names <- factor(c(dfStormFatalitiesTotalOrdered2$event))

barplot(t(as.matrix(dfStormFatalitiesTotalOrdered2)), beside=TRUE)

barplot(dfStormFatalitiesTotalOrdered2) #, dfStormFatalitiesTotalOrdered2$total_fatalities, dfStormFatalitiesTotalOrdered2$event)

# make a barplot which show for entire US the direct fatalities/injuries for each event. The indirect fatalities/injuries are not included in the data  

```

